Tool-driven Revolutions in Archaeological Science

### 2018-09-05 analysis/paper/jcaa-paper.Rmd
more words

Galison's emphasis on tool-driven change, as a complement to Kuhnian concept-driven change, has much potential for archaeology. The centrality of material culture in Galison's view raises the question about whether there is scope for archaeologists to make productive contributions to understanding change in the history science through an archaeological analysis of scientific instrumentation [cf. @Schiffer_2013]. Furthermore, Galison's analysis of the image/logic contrast in the history of physics invites a similar analysis of the history of archaeology Can we identify archaeological traditions focused on data collection and analysis using images and image-making devices [cf. @molyneaux2013cultural], in contrast to digital devices? One candidate for this might be the shift from building relative chronologies based on seriation of visually distinctive artefacts to absolute chronologies based on radiometric dating using computer-controlled instruments. A third issue is how we can use a tool-driven approach not just as a framework for understanding change in science, but as a method to generate change in archaeology?

# Tools Emerging from the Reproducibility Crisis

>>>>

Galison's emphasis on tool-driven change, as a complement to Kuhnian concept-driven change, has much potential for archaeology. The centrality of material culture in Galison's view raises the question about whether there is scope for archaeologists to make productive contributions to understanding change in the history science through an archaeological analysis of scientific instrumentation [cf. @Schiffer_2013]. Furthermore, Galison's analysis of the image/logic contrast in the history of physics invites a similar analysis of the history of archaeology Can we identify archaeological traditions focused on data collection and analysis using images and image-making devices [cf. @molyneaux2013cultural], in contrast to digital devices? One candidate for this might be the shift from building relative chronologies based on seriation of visually distinctive artefacts to absolute chronologies based on radiometric dating using computer-controlled instruments. A third issue is how we can use a tool-driven approach not just as a framework for understanding change in science, but as a method to predict or generate change in archaeology?

# Bibliometric analysis of tool-driven change

The question of predicting or generating change in scientific practice via adoption of new tools is especially intruiging, and in this section we explore how a change in ecology might be relevant for understanding the future of archaeology. @Touchon_McCoy_2016 investigated evaluated changes in statistics used by ecologists as a potential area of tool-driven change. They searched nearly 20,000 published articles in ecology published between 1990 and 2013. They found that there has been a rise in complex and computationally intensive statistical techniques such as mixed effects models and Bayesian statistics and a decline in reliance on approaches such as ANOVA or t-tests. Crucially, they found that ecologists have shifted away from software tools such as SAS and SPSS to the open source program R.

@Touchon_McCoy_2016 identify four factors relating to technological change that might explain the changes they observed in the use of statistics in ecology. First, they note that automated data loggers, GPS trackers, remote sensing, and crowd sourcing have greatly increased rate at which ecologists collect data. Second, increases in desktop and cluster computing power has made complex analytical processes faster and more convienent to compute. Third, the developme of free, open source and easily extensible software such as R allows new methods to spread quickly via online fora and social media

there has been a rise in powerful desktop and supercomputing available to anyone in a university setting. Third, and partly as a result of the increase in computing power available, there has been an explosion in new analytical techniques (e.g., Pinheiro and Bates 2000, Gelman and Hill 2006, McCarthy 2007, Bolker 2008, Legendre and Legendre 2012). The rise of powerful, freely available, open source analytical software like R (R Development Core Team 2013) and associated wiki's, blogs, and listservs where users consult with specialists and find code written by ecologists for ecologists allows the powerful combination of new methods and large data sets to be harnessed by us all. This process has democratized statistical inference, giving the average ecologist the power to conduct analyses previously reserved only for statisticians. Although the analyses that one can conduct in R are often achievable in a program like SAS, the community of ecologist statisticians using R has proliferated rapidly and may be driving change in techniques we document here (although it is admittedly impossible to tease apart cause and effect in this case). This community of users, who share code and advice about analyses, has enabled a great number of ecologists to conduct advanced analyses on their data. Indeed, the flexible nature of statistical programming is what allows those so inclined to “roll their own” statistics. Furthermore, the fact that R and other open source software is free has empowered a global community of users, including those in developing nations that may not have access to commercial costly programs. The fourth and final factor driving the change in statistical practices is that a number of seminal books and papers have fundamentally changed the way many people think about data analyses. To give just one example, Burnham and Anderson's (2002) book on model selection and inference has been cited nearly 30,000 times since its publication in 2002 according to Google Scholar (website accessed 8 December 2015).
